Properties of Winning Iterated Prisoner's Dilemma Strategies.

### 2020-04-17 paper/main.tex
Remove paragraph about TfT
@nikoleta-v3 did some basic analysis (she's going to put the plots up on
the PR) and found that:

- TfT's median C/C_mean was 1.2
- Tf2T's median C/C_mean was 1.37
- Random's median C/C_mean was 0.8

So it's not immediately clear (from the data) that TfT does match the
C_mean (which makes sense if it is never provoked) and in fact, on
average is "as far" from that mean as Random is.

I've added a TODO to include some sort of similar discussion in the
final section.

\begin{abstract}
Researchers have explored the performance of Iterated Prisoner's Dilemma strategies
for decades: from the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated learning
structures such as neural networks, many new strategies have been introduced and tested
in a variety of tournaments and population dynamics. Typical results in the literature,
however, rely on performance against a small number of somewhat arbitrarily selected
strategies in a small number of tournaments, casting doubt on the generalizability
of conclusions. We analyze a large collection of \numberofstrategies
strategies in \numberofalltournaments tournaments, present the top performing strategies across multiple
tournament types, and distill their salient features. The results show that there is not yet a single
strategy that performs well in diverse Iterated Prisoner's Dilemma scenarios.
Nevertheless there are several properties that heavily influence the best performing
strategies, refining the properties described by R. Axelrod in light of
recent and more diverse opponent populations. These are: be nice, be provocable and forgiving,  % TODO modify this based on discussion about vocabulary.
be a little envious, be clever, and adapt to the environment. More precisely,
we find that strategies perform best when their probability of cooperation
matches the total tournament population's aggregate cooperation probabilities,
or a proportion thereof in the case of noisy and probabilistically ending tournaments,
and that the manner in which a strategy achieves the ideal cooperation rate is crucial.
The features of high performing strategies reveal why strategies such as Tit For Tat
performed historically well in tournaments and why zero-determinant strategies
typically do not fare well in tournament settings.
\end{abstract}

@@

\begin{itemize}
    \item \st{Do not be envious} Be a little bit envious
    \item Be ``nice'' in non-noisy environments or when game lengths are longer
    \item Reciprocate both cooperation and defection; Be provocable and forgiving  % TODO Modify this following decision on vocabulary
    \item \st{Do not be too clever} It's ok to be clever
    \item Adapt to the environment; Adjust to the mean population cooperation
\end{itemize}

@@

Moreover, the manner in which a strategy achieves a given cooperation rate
relative to the tournament population average is important. Playing a strategy
that randomly cooperates with probability $C_{\text{mean}}$ is unlikely to be as effective as it will
randomly have cooperations matched with defections, and it may be punished for
defections against cooperations by retailiating opponents. In contrast, TFT naturally
achieves a cooperation rate near $C_{\text{mean}}$ by virtue of copying its opponent's
last move while also minimizing instances where it is exploited by an opponent
(cooperating while the opponent defects), at least in non-noisy tournaments.
This also explains why Tit For N Tats does not fare well for $N > 1$ as it fails
to achieve the proper cooperation ratio by tolerating too many defections.
TFT forces the opponent to pay back the \((C, D)\) round with a \((D, C)\) round before
returning to mutual cooperation. This explains why TFT performed well
in Axelrod's original tournaments as most strategies submitted to those tournaments
were typically cooperative and relatively few strategies played with an easily exploitable
pattern. TFT does not appear in the top ranks in tournaments with more aggressive
strategies because it is too nice. Strategies like Grudger will always
defect after a fixed number of opponent defections, which allows them to
effectively exploit strategies like Alternator or stochastic strategies that
have a non-zero chance of cooperating after mutual defection, which TFT will
not do. Moreover in a noisy environment these strategies will naturally tend
toward always defecting, leading them to exploit strategies like Cooperator. In
such a scenario the noisy environment effectively voids R. Axelrod's rule to be nice,
allowing strategies to attempt exploitation, whereas in a noise-free environment,
attempting exploitation can be risky if strategies that exhibit a Grudger-like behavior
are present, reducing the overall value in attempting to exploit strategies like Cooperator.
% TODO Both NG and MH seem happy with this paragraph but I find it quite a
% mouthful and not very too the point (took me a few reads). If you're both
% happy with it that's fine with me but can we just be sure.

Similarly, these results suggest an explanation regarding the intuitively unexpected
effectiveness of memory-one strategies historically. Given that among the important
features associated with success are the relative cooperation rate to the population
average and the four memory-one probabilities of cooperating conditional on the
previous round of play, these features can be optimized by a memory-one
strategy such as TFT. Usage of more history becomes valuable when there are
exploitable opponent patterns. This is indicated by the importance of SSE as a feature,
showing that the first-approximation provided by a memory-one strategy is no longer
sufficient.

@@

The distributions of SSE and \(CD\) to \(D\) rate for the winners of standard
tournaments are also given in Figure~\ref{fig:discussion_standard}. The SSE
distributions for the winners indicate that the strategy behaved in a ZD way in
several tournaments, however, not constantly. The winners participated in
matches where they did not try to extortionate their opponents. Furthermore, the
\(CD\) to \(D\) distribution indicates that if a strategy were to defect against
the winners they would reciprocate on average with a probability of 0.5.

@@

The distributions of the \(CD\) to \(C\) rate evaluate the behaviour of a
successful strategy after its opponent has defected against it. In standard
tournaments it was observed that a successful strategy reciprocates with a
probability of 0.5. This is distinct between the tournament types. In
tournaments with noise a strategy is less likely to cooperate following a
defection compared to standard tournaments, and in probabilistic ending
tournaments a strategy will reciprocate a defection. In a setting that the type
of the tournament can vary between all the examined types a winning strategy
would reciprocate on average with a probability of 0.58.

@@

This manuscript has explored the performance of \numberofstrategies strategies
of the IPD in a large number of computer tournaments.
We analyzed and extracted the salient features of the best performing strategies
across various tournament types, casting the results in terms of R. Axelrod's
original suggested features of good IPD strategies. Moreover, our results
shed light on the historic performance of TFT, ZDs,
and memory-one strategies generally. Strategies need to match their play
to the cooperativeness of the tournament population and do so in a way that
prevents or minimizes exploitation. Overall we see that complex or clever
strategies can be effective, whether trained against a corpus of possible
opponents or purposely designed to mitigate the impact of noise such as
in the strategy DBS. Further, we showed that while the type of exploitation
attempted by ZDs is not typically effective in tournaments, more
sophisticated strategies capable of selectively exploiting weaker opponents
while mutually cooperating with stronger opponents can be highly successful.
This fact was also indicated numerically by the importance of the strategy
feature SSE in the analysis of strategy features. These results highlight
a central idea in evolutionary game theory in this context: the fitness
landscape is a function of the population (where fitness in this case is
tournament performance). While that may seem obvious now, it shows why
historical tournament results on small or arbitrary populations of strategies
have so often failed to produce generalizable results.

@@

\begin{itemize}
\item be nice, if the environment isn't noisy and the games are long
    \item be provocable and forgiving,
    \item be a little envious,
    \item be clever,
    \item adapt to the environment (including the population of strategies).
\end{itemize}

>>>>

\begin{abstract}
Researchers have explored the performance of Iterated Prisoner's Dilemma strategies
for decades: from the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated learning
structures such as neural networks, many new strategies have been introduced and tested
in a variety of tournaments and population dynamics. Typical results in the literature,
however, rely on performance against a small number of somewhat arbitrarily selected
strategies in a small number of tournaments, casting doubt on the generalizability
of conclusions. We analyze a large collection of \numberofstrategies
strategies in \numberofalltournaments tournaments, present the top performing strategies across multiple
tournament types, and distill their salient features. The results show that there is not yet a single
strategy that performs well in diverse Iterated Prisoner's Dilemma scenarios.
Nevertheless there are several properties that heavily influence the best performing
strategies, refining the properties described by R. Axelrod in light of
recent and more diverse opponent populations. These are: be nice, be somewhat provocable
be a little envious, be clever, and adapt to the environment. More precisely,
we find that strategies perform best when their probability of cooperation
matches the total tournament population's aggregate cooperation probabilities,
or a proportion thereof in the case of noisy and probabilistically ending tournaments,
and that the manner in which a strategy achieves the ideal cooperation rate is crucial.
The features of high performing strategies reveal why strategies such as Tit For Tat
performed historically well in tournaments and why zero-determinant strategies
typically do not fare well in tournament settings.
\end{abstract}

@@

\begin{itemize}
    \item \st{Do not be envious} Be a little bit envious
    \item Be ``nice'' in non-noisy environments or when game lengths are longer
    \item Reciprocate both cooperation and defection correctly; Be somewhat provocable
    \item \st{Do not be too clever} It's ok to be clever
    \item Adapt to the environment; Adjust to the mean population cooperation
\end{itemize}

@@

@@

The distributions of SSE and \(CD\) to \(D\) rate for the winners of standard
tournaments are also given in Figure~\ref{fig:discussion_standard}. The SSE
distributions for the winners indicate that the strategy behaved in a ZD way in
several tournaments, however, not constantly. The winners participated in
matches where they did not try to extortionate their opponents. Furthermore, the
\(CD\) to \(D\) distribution indicates that if a strategy were to defect against
the winners the winners would reciprocate on average with a probability of 0.5.

@@

The distributions of the \(CD\) to \(C\) rate evaluate the behaviour of a
successful strategy after its opponent has defected against it. In standard
tournaments it was observed that a successful strategy reciprocates with a
probability of 0.5.
This leads to the recommendation of needing to be somewhat provocable.
This is distinct between the tournament types. In
tournaments with noise a strategy is less likely to cooperate following a
defection compared to standard tournaments, and in probabilistic ending
tournaments a strategy will reciprocate a defection. In a setting that the type
of the tournament can vary between all the examined types a winning strategy
would reciprocate on average with a probability of 0.58.

@@

This manuscript has explored the performance of \numberofstrategies strategies
of the IPD in a large number of computer tournaments.
We analyzed and extracted the salient features of the best performing strategies
across various tournament types, casting the results in terms of R. Axelrod's
original suggested features of good IPD strategies. Moreover, our results
shed light on the historic performance of TFT, ZDs,
and memory-one strategies generally. Strategies need to match their play
to the cooperativeness of the tournament population and do so in a way that
prevents or minimizes exploitation. Overall we see that complex or clever
strategies can be effective, whether trained against a corpus of possible
opponents or purposely designed to mitigate the impact of noise such as
in the strategy DBS. Further, we showed that while the type of exploitation
attempted by ZDs is not typically effective in tournaments, more
sophisticated strategies capable of selectively exploiting weaker opponents
while mutually cooperating with stronger opponents can be highly successful.
This fact was also indicated numerically by the importance of the strategy
feature SSE in the analysis of strategy features.

% TODO Add discussion about what our results imply and how they are explained
% relative to past performances of TfT and ZDs.

These results highlight
a central idea in evolutionary game theory in this context: the fitness
landscape is a function of the population (where fitness in this case is
tournament performance). While that may seem obvious now, it shows why
historical tournament results on small or arbitrary populations of strategies
have so often failed to produce generalizable results.

@@

\begin{itemize}
\item be nice, if the environment isn't noisy and the games are long
    \item be somewhat provocable,
    \item be a little envious,
    \item be clever,
    \item adapt to the environment (including the population of strategies).
\end{itemize}
